One-arm spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-06 paper/results.tex
update

\section{Results}
We first present results from FARGO simulations. The 2D disc spans
$[R_\mathrm{min}, R_\mathrm{max}] = [0.4,10]R_0$. This gives a total
disc mass $M_{d}=0.086M_*$. The mass within
$R\in[R_\mathrm{min},R_{1}]$ is $0.017M_*$, that within
$R\in[R_{1},R_{2}]$ is $0.049M_*$, and that within
$R\in[R_{2},R_\mathrm{max}]$ is $0.021M_*$. We use a resolution of
$N_R\times N_\phi = 1024\times 2048$, or about $16$ grids per $H$, and
adopt $\epsilon_g=10^{-4}H$ for the   
self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
  approximates for the vertical disc thickness, so a more appropriate
  value would be $\epsilon_g\sim H$ \citep{muller12}. However, because
  $\epsilon_g\propto R$ is needed in FARGO, the Poisson kernel
  (Eq. \ref{2d_grav}) is no longer symmetric in $(R,R^\prime)$. We
  choose a small  
  $\epsilon_g$ in favour of angular momentum conservation, keeping in
  mind that the strength of self-gravity will be over-estimated.}.

In these simulations we subject the disc to initial perturbations in 
cylindrical radial velocity, 
\begin{align}\label{randpert}
  v_R \to v_R+ c_s\frac{\delta}{M}
  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}}{\Delta 
        R}\right)^2\right]}\sum_{m=1}^M\cos{m\phi},
\end{align}
where the amplitude $\delta\in[-10^{-3},10^{-3}]$ is set randomly but
independent of $\phi$, $\overline{R} = (R_{1}+R_{2})/2$
and $\Delta R = (R_{2}-R_{1})/2$.

\subsection{Reference run with random perturbations}
To obtain a picture of the overall disc evolution, here we consider a
fiducial run initialised with $M=10$ in Eq. \ref{randpert}. 
%In this fiducial simulation we subject the disc to initial perturbations in
%cylindrical radial velocity, 
%\begin{align}\label{randpert}
%  v_R \to v_R+ c_s\frac{\delta}{M}
%  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}_d}{\Delta 
%          R_d}\right)^2\right]}\sum_{m=1}^M\cos{m\phi},
%\end{align}
%where $\delta$ is an amplitude, $\overline{R}_d = (R_{1}+R_{2})/2$
%and $\Delta R_d = (R_{2}-R_{1})/2$. We use $M=10$ and choose 
%$\delta\in[-10^{-3},10^{-3}]$ randomly but independent of $\phi$.   
Fig. \ref{fargo_modeamp} plots evolution of the maximum
non-axisymmetric surface density amplitudes in $R\in[R_{1},R_{2}]$
for $m\in[1,10]$. Snapshots from the simulation are shown in
Fig. \ref{fargo_2d}. 
At early times $t\lesssim100P_0$ the disc 
dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
growth initially and saturate (or decays) after $t=40P_0$. Notice the
low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
possibily due to saturation of the high-$m$ modes  
\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
again after $t=70P_0$, and eventually dominates the annulus.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/nonaxi_evol_DZ_fargo}
  \caption{Evolution of maxima in non-axisymmetric surface density 
    in the fiducial FARGO simulation.\label{fargo_modeamp}} 
\end{figure}

\begin{figure}
  \includegraphics[scale=0.27]{figures/polarxy_dens050}\includegraphics[scale=0.27,clip=true,trim=2.26cm 
  0cm 0cm 
  0cm]{figures/polarxy_dens110}\includegraphics[scale=0.27,clip=true,trim=2.26cm
  0cm 0cm 0cm]{figures/polarxy_dens130} 
  \caption{Visualisation of the FARGO 2D simulation. The total  
    non-axisymmetric surface density
    $\Delta\Sigma$ is shown. \label{fargo_2d}} 
\end{figure}

Fig. \ref{2d_angmom} shows the disc angular momenta
evolution. Only the $m=0,\,1$ components are 
plotted since they are dominant. The $m=1$ structure has
an associated negative angular momentum.   
Its growth is compensated by an increase in the axisymmetric
component of angular momentum, such that $\Delta J_0 + \Delta
J_1 \sim 0$. Note that FARGO does not conserve angular momentum
exactly. However, we find the total angular momentum varies by 
$|\Delta J/J|= O(10^{-6})$, and is much smaller in magnitude than the
change in the angular momenta components, $|\Delta J_{0,1}/J|>
O(10^{-5})$. Fig. \ref{2d_angmom} then suggest that angular momentum
is transferred from the one-arm spiral to the background disc.

\begin{figure}
  % scale=0.41
  \includegraphics[width=\linewidth]{figures/nonaxi_evol_ang_fargo}
  \caption{Evolution of angular momentum components in the fiducial
    FARGO simulation. The perturbation
    relative to $t=0$ in 2D is shown in units of the
    initial total angular momentum $J_\mathrm{ref}$.\label{2d_angmom}} 
\end{figure}

\subsection{Trapping the low-frequency $m=1$ spiral}\label{fargo_m1}
In order to focus on the $m=1$ spiral that eventually emerges, here 
we examine a simulation initialised with only $m=1$
perturbations, by setting $M=1$ in Eq. \ref{randpert}. The disc
evolution is largely similar to the previous case, but the final spiral is
more co-herent, facilitating the analysis below. 
Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
distribution at the end of this simulation.

@@

By examining the disc evolution in $t\in[80,120]P_0$, we find the
one-arm spiral in Fig. \ref{2d_fargo_viz} has a 
co-rotation radius and growth rate of
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_k(R_c). 
\end{align*}
The spiral can be considered as low frequency because $|\omega|
\lesssim 0.3\Omega$ (recall $\omega = 
m\Omega_p$) in the region it is promininent $R\in[R_{1},R_{2}]$. The
growth rate is also slow relative to the local rotation. This gives a
nearly stationary spiral pattern.

%Since most of
%the perturbation lies at $R<R_c$, this mode has negative
%angular momentum according to local linear theory.

Next, we write $\Sigma_1 = 
|\Sigma_1|\exp{(\ii kR)}$, where $k$ is real, and assume the amplitude
$|\Sigma_1|$ varies slowly compared to the complex phase. This is the
main assumption in local theory. We plot in
Fig. \ref{fargo_wavenumber} the normalise wavenumber associated with
the $m=1$ spiral. We find that
\begin{align*}
  kR \sim \frac{\pi G \Sigma}{c_s^2}R \sim \frac{1}{hQ}, 
\end{align*}
where we used $Q\sim c_s\Omega/\pi G \Sigma$ and $R\Omega/c_s\sim
h^{-1}$. Since $Q=O(1)$ and $h\ll 1$, $|kR|\gg 1$ and we 
can apply results from local theory self-consistently. Since $k>0$,
the spiral is trailing, as seen in Fig. \ref{2d_fargo_viz}.

@@

Using the estimated value of $R_c$, we plot in
Fig. \ref{fargo_qbarrier} the quantity $\nu^2 - 1 + Q^{-2}$, which is
required to be positive in local theory for purely wave-like
solutions to the dispersion relation (Eq. \ref{dispersion}) when the
mode frequency is given. The figure shows two $Q$-barriers located in
the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$, which is indeed the
region where the $m=1$ spiral develops. This means that the one-arm
spiral is trapped. There is one outer Lindblad resonance at $R_L\simeq
7.2R_0$. Thus, waves may be launched in $R\gtrsim 7.2R_0$ by the main 
spiral disturbance in the inner disc. %this would be a de-stabilising
% effect.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_Qbar_fargo.ps} 
  \caption{Dimensionless mode frequency $\nu$ for the $m=1$ spiral in
    Fig. \ref{2d_fargo_viz}. The disperion relation for local density
    waves, Eq. \ref{dispersion}, permits purely wave-like solutions in 
    regions where $\nu^2 - 1 + Q^{-2}>0$. 
    \label{fargo_qbarrier}} 
\end{figure}

@@

We plot growth rates of the $m=1$ spiral as a function of $q$ in 
Fig. \ref{fargo_varq_growth}. The correlation is well-fitted by a
linear relation
\begin{align*}
  \gamma \simeq 0.015 q \Omega_k(R_0). 
\end{align*}
This confirms that the temperature gradient is responsible for the
development of the one-arm spiral observed in our simulations. 
\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varq.ps}    
  \caption{Growth rates of the $m=1$ spiral mode as a function of the
    imposed sound-speed gradient $q$ (asterisks). A linear fit is also
    plotted (dotted line). 
    \label{fargo_varq_growth}} 
\end{figure}

@@

Fig. \ref{fargo_varh_growth} shows the growth rates of the $m=1$
spiral in $R\in[R_1,R_2]$ as a function of $h$. Growth rates increases
with $h$, roughly as 
\begin{align}
  \gamma \simeq 0.10h\Omega_k(R_0).  
\end{align}
However, the linear fit is less good than for variable $q$ above. This
may be due to the change in the total disc mass when $h$ changes. We
find no qualitative difference between the spiral pattern that
emerges.

>>>>

\section{Results}
We first present results from FARGO simulations. The 2D disc spans
$[R_\mathrm{min}, R_\mathrm{max}] = [0.4,10]R_0$. This gives a total
disc mass $M_{d}=0.086M_*$. The mass within
$R\in[R_\mathrm{min},R_{1}]$ is $0.017M_*$, that within
$R\in[R_{1},R_{2}]$ is $0.049M_*$, and that within
$R\in[R_{2},R_\mathrm{max}]$ is $0.021M_*$. We use a resolution of
$N_R\times N_\phi = 1024\times 2048$, or about $16$ grids per $H$, and
adopt $\epsilon_g=10^{-4}hR$ for the   
self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
  approximates for the vertical disc thickness, so a more appropriate
  value would be $\epsilon_g\sim H$ \citep{muller12}. However, because
  $\epsilon_g\propto R$ is needed in FARGO, the Poisson kernel
  (Eq. \ref{2d_grav}) is no longer symmetric in $(R,R^\prime)$. We
  choose a small  
  $\epsilon_g$ in favour of angular momentum conservation, keeping in
  mind that the strength of self-gravity will be over-estimated.}.

In these simulations the disc is subject to initial perturbations in 
cylindrical radial velocity, 
\begin{align}\label{randpert}
  v_R \to v_R+ c_s\frac{\delta}{M}
  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}}{\Delta 
          R}\right)^2\right]}\sum_{m=1}^M\cos{m\phi},
\end{align}
where the amplitude $\delta\in[-10^{-3},10^{-3}]$ is set randomly but
independent of $\phi$, $\overline{R} = (R_{1}+R_{2})/2$
and $\Delta R = (R_{2}-R_{1})/2$.

\subsection{Reference run with random perturbations}
To obtain a picture of the overall disc evolution, we describe a 
fiducial run initialised with $M=10$ in Eq. \ref{randpert}. 
Fig. \ref{fargo_modeamp} plots evolution of the maximum 
non-axisymmetric surface density amplitudes in $R\in[R_{1},R_{2}]$
for $m\in[1,10]$. Snapshots from the simulation are shown in
Fig. \ref{fargo_2d}. 
At early times $t\lesssim100P_0$ the disc 
dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
growth initially and saturate (or decays) after $t=40P_0$. Notice the
low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
possibily due to saturation of the high-$m$ modes  
\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
again after $t=70P_0$, and eventually dominates the annulus.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/nonaxi_evol_DZ_fargo}
  \caption{Evolution of maxima in non-axisymmetric surface density 
    in the fiducial FARGO simulation initialised with perturbations
    with $m\in[1,10]$.\label{fargo_modeamp}} 
\end{figure}

\begin{figure}
  \includegraphics[scale=0.27]{figures/polarxy_dens050}\includegraphics[scale=0.27,clip=true,trim=2.26cm 
  0cm 0cm 
  0cm]{figures/polarxy_dens110}\includegraphics[scale=0.27,clip=true,trim=2.26cm
  0cm 0cm 0cm]{figures/polarxy_dens130} 
  \caption{Visualisation of the fiducial FARGO 2D simulation in
    Fig. \ref{fargo_modeamp}. The total  
    non-axisymmetric surface density
    $\Delta\Sigma$ is shown. \label{fargo_2d}} 
\end{figure}

Fig. \ref{2d_angmom} shows the evolution of disc angular momentum
components. Only the $m=0,\,1$ components are 
plotted since they are dominant. The $m=1$ structure has
an associated negative angular momentum.  
Its growth is compensated by an increase in the axisymmetric
component of angular momentum, such that $\Delta J_0 + \Delta
J_1 \sim 0$. Note that FARGO does not conserve angular momentum
exactly. However, we find the total angular momentum varies by 
$|\Delta J/J|= O(10^{-6})$, and is much smaller in magnitude than the
change in the angular momenta components, $|\Delta J_{0,1}/J|>
O(10^{-5})$. Fig. \ref{2d_angmom} then suggest that angular momentum
is transferred from the one-arm spiral to the background disc.

\begin{figure}
  % scale=0.41
  \includegraphics[width=\linewidth]{figures/nonaxi_evol_ang_fargo}
  \caption{Evolution of angular momentum components in the fiducial
    FARGO simulation in Fig. \ref{fargo_modeamp}---\ref{fargo_2d}. The
    perturbation relative to $t=0$ in 2D is shown in units of the 
    initial total angular momentum $J_\mathrm{ref}$.\label{2d_angmom}} 
\end{figure}

\subsection{Trapping the low-frequency $m=1$ spiral}\label{fargo_m1}
In order to focus on the $m=1$ spiral that eventually emerges, here 
we examine a simulation initialised with only $m=1$
perturbations, by setting $M=1$ in Eq. \ref{randpert}. The disc
evolution is largely similar to the previous case, but the final spiral is
more coherent, which facilitates the analysis below.

Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
at the end of this simulation. For this one-arm spiral
we measure a co-rotation radius and growth rate of 
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_k(R_c). 
\end{align*}
The spiral can be considered as low frequency because its 
pattern speed $\Omega_p\lesssim 0.3\Omega$ in $R\in[R_{1},R_{2}]$,
where it is  most prominent. The growth rate $\gamma$ is also slow
relative to the local rotation. Thus, the spiral 
pattern is nearly stationary.

@@

Next, we write $\Sigma_1 = 
|\Sigma_1|\exp{(\ii kR)}$, where $k$ is real, and assume the amplitude
$|\Sigma_1|$ varies slowly compared to the complex phase. This is the
main assumption in local theory. We plot in
Fig. \ref{fargo_wavenumber} the normalised radial wavenumber associated with
the $m=1$ spiral. We find 
\begin{align*}
  kR \sim \frac{\pi G \Sigma}{c_s^2}R \sim \frac{1}{hQ}, 
\end{align*}
where we used $Q\sim c_s\Omega/\pi G \Sigma$ and $R\Omega/c_s\sim
h^{-1}$. Since $Q=O(1)$ and $h\ll 1$ imply $|kR|\gg 1$, we 
can apply results from local theory self-consistently. Recall $k>0$
means the spiral is trailing, as seen in Fig. \ref{2d_fargo_viz}.

@@

%%%%%%%%%%%%%%%%%%%%%%

Using the estimated value of $R_c$, we plot in
Fig. \ref{fargo_qbarrier} the quantity $\nu^2 - 1 + Q^{-2}$, which is
required to be positive in local theory for purely wave-like
solutions to the dispersion relation (Eq. \ref{dispersion}) when the
mode frequency is given. The figure shows two $Q$-barriers located in
the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$, which is indeed the
region where the $m=1$ spiral develops. This suggests that the one-arm
spiral is trapped. There is one outer Lindblad resonance at $R_L\simeq
7.2R_0$. Thus, waves may be launched in $R\gtrsim 7.2R_0$ by the main 
spiral disturbance in the inner disc \citep{lin11b}. 
%this would be a de-stabilising
% effect.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_Qbar_fargo.ps} 
  \caption{Dimensionless mode frequency $\nu$ for the $m=1$ spiral in
    Fig. \ref{2d_fargo_viz}. For a given real mode frequency, the
    dispersion relation for local density waves, Eq. \ref{dispersion},
    permits purely wave-like solutions in regions where $\nu^2 - 1 +
    Q^{-2}>0$.    
    \label{fargo_qbarrier}} 
\end{figure}

@@

We plot growth rates of the $m=1$ spiral as a function of $q$ in 
Fig. \ref{fargo_varq_growth}. The correlation is well-fitted by a
linear relation
\begin{align*}
  \gamma \simeq \left[0.015 q - 7.9\times10^{-4}\right] \Omega_k(R_0). 
\end{align*}
This confirms that the temperature gradient is responsible for the
development of the one-arm spiral observed in our simulations. 
\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varq.ps}    
  \caption{Growth rates of the $m=1$ spiral mode as a function of the
    imposed sound-speed gradient $q$ (asterisks). A linear fit is also
    plotted (dotted line). 
    \label{fargo_varq_growth}} 
\end{figure}

@@

Fig. \ref{fargo_varh_growth} shows the growth rates of the $m=1$
spiral in $R\in[R_1,R_2]$ as a function of $h$. Growth rates increases
with $h$, roughly as 
\begin{align}
  \gamma \simeq \left[0.10h + 8.3\times10^{-3}\right]\Omega_k(R_0).  
\end{align}
However, the linear fit is less good than for variable $q$ above. This
may be due to the change in the total disc mass when $h$ changes. We
find no qualitative difference between the spiral pattern that
emerges.
